The Design and Implementation of the Synchronous Language \CEU

### 2016-03-29 paper.tex
...

\usepackage{verbatim}
\usepackage{xspace}
\newcommand{\C}{\emph{C}\xspace}
\newcommand{\VM}{\emph{VM}\xspace}
\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
\newcommand{\code}[1] {{\small{\texttt{#1}}}}
\newcommand{\MM}[1] {\textcircled{\tiny{\textsf{#1}}}}

@@

Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel (a) and 
\CEU (b) for the following control specification:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''}~\cite{esterel.primer}.
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 3--8, in both 
implementations), i.e., Esterel's `$\|$' and \CEU's \code{par/and} constructs 
are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 2--9), which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 2--11):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

.
\end{lstlisting}
\centering\small{(a) Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
loop do
   par/or do
      par/and do
         await A;
      with
         await B;
      end
      emit O;
   with
      await R;
   end
end
\end{lstlisting}
\centering\small{(b) \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ A control specification implemented in Esterel and \CEU:
\emph{``Emit \code{O} after \code{A} and \code{B}, resetting each \code{R}.''}
\label{lst.abro}
}
\end{figure}

@@

Esterel makes no semantic distinctions between internal and external signals.
In particular, programs can emit external and internal signals simultaneously, 
with all coexisting during the entire reaction.
%
%In \CEU, a reaction starts from an external input event and programs cannot 
%emit inputs at all.
%Therefore, the occurring input event is unique during the entire reaction, 
%resulting in intrinsic queue-based handling.
%
In \CEU, subsequent external events coming from the environment define a 
timeline, which programs cannot interfere by emitting new external events.
%
To cope with intra-reaction communication, \CEU supports internal events which 
programs can \code{emit} and \code{await}.
In contrast with external events, internal events follow a stack-based 
execution policy similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel (a) and \CEU (b).
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln. 4--5) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
in no particular order.
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (ln. 4), emits \code{b}, and pauses.
\item 2nd trail awakes (ln. 8), calls \code{\_g()}, and terminates.
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

\begin{figure}[b]
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\centering\small{(a) Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\centering\small{(b) \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU: similar syntax, but 
different semantics.
\label{lst.prints}
}
\end{figure}

@@

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\centering\small{(b)}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
Example (a) is safe because the trails access \code{x} atomically in different 
reactions;
Example (b) is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure}

Even so, we consider that enforcing an arbitrary execution order can be 
misleading in some cases.
%
For instance, consider the two examples in Figure~\ref{lst.shared}, both of 
which define a shared variable (ln. 2), and assign to it in parallel trails 
(ln.  5, 8).
%
In the example \code{(a)}, the two assignments to \code{x} execute from 
reactions to different events \code{A} and \code{B} which, by definition, 
cannot occur simultaneously (Section~\ref{sec.ceu.exts}).
Hence, for the sequence \code{A->B}, \code{x} becomes \code{(1+1)*2->4}, while 
for \code{B->A}, \code{x} becomes \code{(1*2)+1->3}.
%
In the example \code{(b)}, the two assignments to \code{y} are simultaneous 
because they execute in reaction to the same event \code{A}.
Since \CEU employs lexical order for intra-reaction statements, the execution 
is still deterministic, and \code{y} always becomes \code{(1+1)*2->4}.
%
However, an (apparently innocuous) change in the order of trails changes the 
semantics of the program, which we consider unsafe.

To mitigate this threat, \CEU performs a temporal analysis at compile time and 
detects concurrent accesses to shared variables, as follows:
if a variable is written in a trail segment, then a concurrent trail segment 
cannot read or write to that variable, nor dereference a pointer of that 
variable type.
%An analogous policy is applied for pointers \emph{vs} variables and pointers 
%\emph{vs} pointers.
%
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
Concurrency in \CEU is characterized when two or more trail segments in 
parallel react to the same input event.
%
Considering the examples in Figure~\ref{lst.shared}:
%
\begin{itemize}
\item The assignments to \code{x} in lines 2 and 5 in \code{(a)}
      \textbf{cannot} be concurrent because they are not in parallel trails.
\item The assignments to \code{x} in lines 5 and 8 in \code{(a)}
      \textbf{cannot} be concurrent because they \textbf{cannot} execute during 
      the same reaction.
\item The assignments to \code{y} in lines 5 and 8 in \code{(b)}
      \textbf{can} be concurrent because they are in parallel trails and 
      \textbf{can} execute during the same reaction.
\end{itemize}
%
The algorithm for the analysis, which is depicted in 
Section~\ref{sec.impl.analysis}, inspects all possible \code{await} statements 
that precede a variable access and keep a list with all corresponding awaking 
events.
Then, it checks all accesses in parallel trails to see if they share an awaking 
event.
If it is the case, the compiler warns about the suspicious accesses.

@@

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
native do
    #define NUM 10
    void f  (void)  { <...> }
    void g  (int v) { <...> }
    int  id (int v) { <...> }
end
par/and do
    _f();
with
    _g(_id(_NUM));
end
\end{lstlisting}
\centering\small{(a) Definition and use of external symbols}
\end{minipage}
%
\begin{minipage}[t]{0.44\linewidth}
\begin{lstlisting}
native @const _NUM;
native @pure  _id;
native @safe  _f with _g;

.
\end{lstlisting}
\centering\small{(b) Annotations for the external symbols in (a).}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ The unsafe program in (a) only compiles with the annotations in (b).
\label{lst.c}
}
\end{figure}

@@

The code in Figure~\ref{lst.c}.b uses annotations to provide hints to the 
compiler about the semantics of the \C symbols in use in \code{(a)}:
%
\begin{itemize}
\item \code{\_NUM} is a constant symbol, meaning that it is safe to use it 
      concurrently with any other symbol in the program.
\item \code{\_id} is a pure function, also meaning that it is safe to call it
      concurrently with any other symbol in the program.
\item Both \code{\_f} and \code{\_g} are impure, but their effects do not 
      conflict and they can be safely called concurrently.
\end{itemize}

With the annotations in \code{(b)}, the program in \code{(a)} compiles 
successfully.

@@

The example in Figure~\ref{lst.fin} in Esterel (\code{(a)}) and \CEU 
(\code{(b)}) uses the \code{lock} and \code{unlock} calls which represent 
accesses to an external resource.
The normal behavior is to \code{lock} the resource (ln. \code{(a)}:4 and 
\code{(b)}:3), perform some operations in subsequent reactions to input 
\code{A} (ln. \code{(a)}:5--8 and \code{(b)}:7--10), and then \code{unlock} the 
resource (ln. \code{(a)}:10 and \code{(b)}:5).
However, if the aborting input \code{B} (ln. \code{(a)}:12 and \code{(b)}:12) 
occurs after the \code{lock} but before the reactions to \code{A}, we still 
want to call \code{unlock} to safely release the resource.
%
In Esterel and \CEU, the finalize clauses (ln. \code{(a)}:10 and \code{(b)}:5) 
are automatically called if the enclosing blocks (ln. \code{(a)}:3--1 and 
\code{(b)}:3--10) are externally aborted (ln. \code{(a)}:12 and \code{(b)}:12).

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
input A, B;
abort
    finalize
        call lock();
        await A;
        <...>;      // do something
        await A;
        <...>;      // do something
    with
        call unlock();
    end
when B
.
\end{lstlisting}
\centering\small{(a) Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
input void A, B;
par/or do
    _lock();
    finalize with
        _unlock();
    end
    await A;
    <...>;      // do something
    await A;
    <...>;      // do something
with
    await B;
end
\end{lstlisting}
\centering\small{(b) \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{
Finalization in Esterel and \CEU: after the call to \code{lock}, both languages 
guarantee to call \code{unlock} if the enclosing block aborts when \code{B} 
occurs.
\label{lst.fin}
}
\end{figure}

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
par/or do
   var _buffer_t msg;
   <...> // prepare msg
   finalize
      _send_request(&msg);
   with
      _send_cancel(&msg);
   end
   await SEND_ACK;
with
   <...>
end
.
\end{lstlisting}
\centering\small{(a) Local resource finalization}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
par/or do
   var _FILE* f;
   finalize
      f = _fopen(...);
   with
      _fclose(f);
   end
   _fwrite(..., f);
   await A;
   _fwrite(..., f);
with
   <...>
end
\end{lstlisting}
\centering\small{(b) External resource finalization}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{
\CEU enforces the use of finalization to prevent \emph{dangling pointers} for 
local resources and \emph{memory leaks} for external resources.
\label{lst.fin.ceu}
}
\end{figure}

@@

.
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.51\linewidth}
\begin{lstlisting}
par/or do
    await 10ms;
    <...>         // any non-awaiting sequence
    await  1ms;
    v = 1;
with
    await 12ms;
    v = 2;
end
\end{lstlisting}
\centering\small{(b)}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{ First-class timers in \CEU.
\label{lst.timers}
}
\end{figure}

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
input void A, B;
var int y;
par/or do
  await A;
  y = 1;
with
  await B;
  y = 2;
end
await A;
y = 3;
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
Stmts I={.} O={A}
    Dcl_y I={.} O={.}
    ParOr I={.} O={A,B}
        Stmts I={.} O={A}
            Await_A I={.} O={A}
            Set_y I={A} O={A}
        Stmts I={.} O={B}
            Await_B I={.} O={B}
            Set_y I={B} O={B}
    Await_A I={A,B} O={A}
    Set_y I={A} O={A}
\end{lstlisting}
\centering\small{(b)}
\end{minipage}
%
%\rule{14cm}{0.37pt}
\caption{ A program with a corresponding AST describing the sets $I$ and $O$.
%{\small %\textmd{
The program is safe because accesses to \code{y} in parallel have no 
intersections for $I$.
%}%}
\label{lst.impl.ast}
}
\end{figure}

@@

In the final generated code in \C, each trail segment label representing an 
entry point becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in (a), the compiler extracts the entry points and associated 
trails, e.g., the label \code{Awake\_e} will execute on \code{TRAIL-0} (ln. 7).
%
For each statement that pauses (\code{emit} and \code{await}), resumes 
(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
and \code{break}), the compiler splits the trail into segments with associated 
entry points.
%
The entry points translate to an \code{enum} in the generated code (ln. 1--10, 
in (b)).
The state of trails translates to a vector of type \code{trail\_t} with the 
maximum number of simultaneous trails (ln. 12--15).
On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
(ln. 13), while all others are set to \code{INACTIVE} (in the example, only 
one, in ln. 14).

@@

.
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.31\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
enum {
  Main = 1,  // ln 3
  Awake_e,   // ln 7
  And_chk,   // ln 8,15
  And_sub_2, // ln 10
  Awake_A_1, // ln 12
  Emit_cont, // ln 14
  And_out,   // ln 17
  Awake_A_2  // ln 19
};

@@

.
\end{lstlisting}
\centering\small{(b)}
\end{minipage}
%
\begin{minipage}[t]{0.37\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
void dispatch (trail_t* t) {
  switch (t->lbl) {
    case Main:
      // activate TRAIL 1
      TRLS[1].evt = STACKED;
      TRLS[1].lbl = And_sub_2;
      TRLS[1].stk = cur_stack;

@@

    <...>  // other labels
  }
}
\end{lstlisting}
\centering\small{(c)}
\end{minipage}
\caption{
%{\small %\textmd{
(a) Static allocation of trails: the comments identify the trail indexes 
inferred by the compiler;
(b) Entry-point labels: each trail segment has an associated numeric identifier 
generated by the compiler.
(c) Dispatch function: uses a \code{switch} to associate each segment 
identifier with the corresponding code to execute.
%}%}
\label{lst.impl.trails}
}
\end{figure}

@@

.
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.49\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
// Output events
void VM.out(int evt_id, void* args) {
    switch (id){
        case O_REQUEST_TEMPERATURE:
            call TINYOS_TEMP.read();
        <...>; // O_REQUEST_SEND
    }
}

@@

// System calls
void VM.function(int id, void* params) {
    switch (id) {
        case F_GET_RADIO_ID:
            VM.push(TINYOS_NODE_ID);
    }
}
\end{lstlisting}
\centering\small{(b)}
\end{minipage}
%
\caption{
%{\small %\textmd{
(a) \CEU interface with customized \VM.
%
(b) The routine \code{VM.out} redirects all output events to the corresponding 
OS calls (ln.  1--8).
%
Each \emph{TinyOS} event callback calls \code{VM.enqueue} for the corresponding 
input event (ln 10--14).
%
System calls use \code{VM.push} for immediate return values (ln. 16--22).
%
\label{lst.terra.defs}
}
\end{figure}

>>>>

\usepackage{verbatim}
\usepackage{xspace}
\newcommand{\C}{\emph{C}\xspace}
\newcommand{\VM}{\emph{VM}\xspace}
\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
\newcommand{\code}[1] {{\small{\texttt{#1}}}}
\newcommand{\ax}{\code{[a]}\xspace}
\newcommand{\bx}{\code{[b]}\xspace}
\newcommand{\MM}[1] {\textcircled{\tiny{\textsf{#1}}}}

@@

Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel \ax and 
\CEU \bx for the following control specification:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''}~\cite{esterel.primer}.
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 3--8, in both 
implementations), i.e., Esterel's `$\|$' and \CEU's \code{par/and} constructs 
are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 2--9), which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 2--11):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

.
\end{lstlisting}
\centering\small{\ax Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
loop do
   par/or do
      par/and do
         await A;
      with
         await B;
      end
      emit O;
   with
      await R;
   end
end
\end{lstlisting}
\centering\small{\bx \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ A control specification implemented in Esterel and \CEU:
\emph{``Emit \code{O} after \code{A} and \code{B}, resetting each \code{R}.''}
\label{lst.abro}
}
\end{figure}

@@

Esterel makes no semantic distinctions between internal and external signals.
In particular, programs can emit external and internal signals simultaneously, 
with all coexisting during the entire reaction.
%
%In \CEU, a reaction starts from an external input event and programs cannot 
%emit inputs at all.
%Therefore, the occurring input event is unique during the entire reaction, 
%resulting in intrinsic queue-based handling.
%
In \CEU, subsequent external events coming from the environment define a 
timeline, which programs cannot interfere by emitting new external events.
%
To cope with intra-reaction communication, \CEU supports internal events which 
programs can \code{emit} and \code{await}.
In contrast with external events, internal events follow a stack-based 
execution policy similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel \ax and \CEU \bx.
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln. 4--5) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
in no particular order.
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (ln. 4), emits \code{b}, and pauses.
\item 2nd trail awakes (ln. 8), calls \code{\_g()}, and terminates.
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

\begin{figure}[b]
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\centering\small{\ax Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\centering\small{\bx \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU: similar syntax, but 
different semantics.
\label{lst.prints}
}
\end{figure}

@@

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\centering\small{\ax}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\centering\small{\bx}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
Example \ax is safe because the trails access \code{x} atomically in different 
reactions;
Example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure}

Even so, we consider that enforcing an arbitrary execution order can be 
misleading in some cases.
%
For instance, consider the two examples in Figure~\ref{lst.shared}, both of 
which define a shared variable (ln. 2), and assign to it in parallel trails 
(ln.  5, 8).
%
In the example \ax, the two assignments to \code{x} execute from reactions to 
different events \code{A} and \code{B} which, by definition, cannot occur 
simultaneously (Section~\ref{sec.ceu.exts}).
Hence, for the sequence \code{A->B}, \code{x} becomes \code{(1+1)*2->4}, while 
for \code{B->A}, \code{x} becomes \code{(1*2)+1->3}.
%
In the example \bx, the two assignments to \code{y} are simultaneous because 
they execute in reaction to the same event \code{A}.
Since \CEU employs lexical order for intra-reaction statements, the execution 
is still deterministic, and \code{y} always becomes \code{(1+1)*2->4}.
%
However, an (apparently innocuous) change in the order of trails changes the 
semantics of the program, which we consider unsafe.

To mitigate this threat, \CEU performs a temporal analysis at compile time and 
detects concurrent accesses to shared variables, as follows:
if a variable is written in a trail segment, then a concurrent trail segment 
cannot read or write to that variable, nor dereference a pointer of that 
variable type.
%An analogous policy is applied for pointers \emph{vs} variables and pointers 
%\emph{vs} pointers.
%
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
Concurrency in \CEU is characterized when two or more trail segments in 
parallel react to the same input event.
%
Considering the examples in Figure~\ref{lst.shared}:
%
\begin{itemize}
\item The assignments to \code{x} in lines 2 and 5 in \ax
      \textbf{cannot} be concurrent because they are not in parallel trails.
\item The assignments to \code{x} in lines 5 and 8 in \ax
      \textbf{cannot} be concurrent because they \textbf{cannot} execute during 
      the same reaction.
\item The assignments to \code{y} in lines 5 and 8 in \bx
      \textbf{can} be concurrent because they are in parallel trails and 
      \textbf{can} execute during the same reaction.
\end{itemize}
%
The algorithm for the analysis, which is depicted in 
Section~\ref{sec.impl.analysis}, inspects all possible \code{await} statements 
that precede a variable access and keep a list with all corresponding awaking 
events.
Then, it checks all accesses in parallel trails to see if they share an awaking 
event.
If it is the case, the compiler warns about the suspicious accesses.

@@

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
native do
    #define NUM 10
    void f  (void)  { <...> }
    void g  (int v) { <...> }
    int  id (int v) { <...> }
end
par/and do
    _f();
with
    _g(_id(_NUM));
end
\end{lstlisting}
\centering\small{\ax Definition and use of external symbols}
\end{minipage}
%
\begin{minipage}[t]{0.44\linewidth}
\begin{lstlisting}
native @const _NUM;
native @pure  _id;
native @safe  _f with _g;

.
\end{lstlisting}
\centering\small{\bx Annotations for the external symbols in \ax.}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ The unsafe program in \ax only compiles with the annotations in \bx.
\label{lst.c}
}
\end{figure}

@@

The code in Figure~\ref{lst.c}.b uses annotations to provide hints to the 
compiler about the semantics of the \C symbols in use in \ax:
%
\begin{itemize}
\item \code{\_NUM} is a constant symbol, meaning that it is safe to use it 
      concurrently with any other symbol in the program.
\item \code{\_id} is a pure function, also meaning that it is safe to call it
      concurrently with any other symbol in the program.
\item Both \code{\_f} and \code{\_g} are impure, but their effects do not 
      conflict and they can be safely called concurrently.
\end{itemize}

With the annotations in \bx, the program in \ax compiles successfully.

@@

The example in Figure~\ref{lst.fin} in Esterel \ax and \CEU \bx uses the 
\code{lock} and \code{unlock} calls which represent accesses to an external 
resource.
The normal behavior is to \code{lock} the resource (ln. \ax:4 and \bx:3), 
perform some operations in subsequent reactions to input \code{A} (ln.  
\ax:5--8 and \bx:7--10), and then \code{unlock} the resource (ln.  \ax:10 and 
\bx:5).
However, if the aborting input \code{B} (ln. \ax:12 and \bx:12) occurs after 
the \code{lock} but before the reactions to \code{A}, we still want to call 
\code{unlock} to safely release the resource.
%
In Esterel and \CEU, the finalize clauses (ln. \ax:10 and \bx:5) are 
automatically called if the enclosing blocks (ln. \ax:3--1 and \bx:3--10) are 
externally aborted (ln. \ax:12 and \bx:12).

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
input A, B;
abort
    finalize
        call lock();
        await A;
        <...>;      // do something
        await A;
        <...>;      // do something
    with
        call unlock();
    end
when B
.
\end{lstlisting}
\centering\small{\ax Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
input void A, B;
par/or do
    _lock();
    finalize with
        _unlock();
    end
    await A;
    <...>;      // do something
    await A;
    <...>;      // do something
with
    await B;
end
\end{lstlisting}
\centering\small{\bx \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{
Finalization in Esterel and \CEU: after the call to \code{lock}, both languages 
guarantee to call \code{unlock} if the enclosing block aborts when \code{B} 
occurs.
\label{lst.fin}
}
\end{figure}

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
par/or do
   var _buffer_t msg;
   <...> // prepare msg
   finalize
      _send_request(&msg);
   with
      _send_cancel(&msg);
   end
   await SEND_ACK;
with
   <...>
end
.
\end{lstlisting}
\centering\small{\ax Local resource finalization}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
par/or do
   var _FILE* f;
   finalize
      f = _fopen(...);
   with
      _fclose(f);
   end
   _fwrite(..., f);
   await A;
   _fwrite(..., f);
with
   <...>
end
\end{lstlisting}
\centering\small{\bx External resource finalization}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{
\CEU enforces the use of finalization to prevent \emph{dangling pointers} for 
local resources and \emph{memory leaks} for external resources.
\label{lst.fin.ceu}
}
\end{figure}

@@

.
\end{lstlisting}
\centering\small{\ax}
\end{minipage}
%
\begin{minipage}[t]{0.51\linewidth}
\begin{lstlisting}
par/or do
    await 10ms;
    <...>         // any non-awaiting sequence
    await  1ms;
    v = 1;
with
    await 12ms;
    v = 2;
end
\end{lstlisting}
\centering\small{\bx}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{ First-class timers in \CEU.
\label{lst.timers}
}
\end{figure}

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
input void A, B;
var int y;
par/or do
  await A;
  y = 1;
with
  await B;
  y = 2;
end
await A;
y = 3;
\end{lstlisting}
\centering\small{\ax}
\end{minipage}
%
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
Stmts I={.} O={A}
    Dcl_y I={.} O={.}
    ParOr I={.} O={A,B}
        Stmts I={.} O={A}
            Await_A I={.} O={A}
            Set_y I={A} O={A}
        Stmts I={.} O={B}
            Await_B I={.} O={B}
            Set_y I={B} O={B}
    Await_A I={A,B} O={A}
    Set_y I={A} O={A}
\end{lstlisting}
\centering\small{\bx}
\end{minipage}
%
%\rule{14cm}{0.37pt}
\caption{ A program with a corresponding AST describing the sets $I$ and $O$.
%{\small %\textmd{
The program is safe because accesses to \code{y} in parallel have no 
intersections for $I$.
%}%}
\label{lst.impl.ast}
}
\end{figure}

@@

In the final generated code in \C, each trail segment label representing an 
entry point becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in \ax, the compiler extracts the entry points and associated 
trails, e.g., the label \code{Awake\_e} will execute on \code{TRAIL-0} (ln. 7).
%
For each statement that pauses (\code{emit} and \code{await}), resumes 
(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
and \code{break}), the compiler splits the trail into segments with associated 
entry points.
%
The entry points translate to an \code{enum} in the generated code (ln. 1--10, 
in \bx).
The state of trails translates to a vector of type \code{trail\_t} with the 
maximum number of simultaneous trails (ln. 12--15).
On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
(ln. 13), while all others are set to \code{INACTIVE} (in the example, only 
one, in ln. 14).

@@

.
\end{lstlisting}
\centering\small{\ax}
\end{minipage}
%
\begin{minipage}[t]{0.31\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
enum {
  Main = 1,  // ln 3
  Awake_e,   // ln 7
  And_chk,   // ln 8,15
  And_sub_2, // ln 10
  Awake_A_1, // ln 12
  Emit_cont, // ln 14
  And_out,   // ln 17
  Awake_A_2  // ln 19
};

@@

.
\end{lstlisting}
\centering\small{\bx}
\end{minipage}
%
\begin{minipage}[t]{0.37\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
void dispatch (trail_t* t) {
  switch (t->lbl) {
    case Main:
      // activate TRAIL 1
      TRLS[1].evt = STACKED;
      TRLS[1].lbl = And_sub_2;
      TRLS[1].stk = cur_stack;

@@

    <...>  // other labels
  }
}
\end{lstlisting}
\centering\small{(c)}
\end{minipage}
\caption{
%{\small %\textmd{
\ax Static allocation of trails: the comments identify the trail indexes 
inferred by the compiler;
\bx Entry-point labels: each trail segment has an associated numeric identifier 
generated by the compiler.
(c) Dispatch function: uses a \code{switch} to associate each segment 
identifier with the corresponding code to execute.
%}%}
\label{lst.impl.trails}
}
\end{figure}

@@

.
\end{lstlisting}
\centering\small{\ax}
\end{minipage}
%
\begin{minipage}[t]{0.49\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
// Output events
void VM.out(int evt_id, void* args) {
    switch (id){
        case O_REQUEST_TEMPERATURE:
            call TINYOS_TEMP.read();
        <...>; // O_REQUEST_SEND
    }
}

@@

// System calls
void VM.function(int id, void* params) {
    switch (id) {
        case F_GET_RADIO_ID:
            VM.push(TINYOS_NODE_ID);
    }
}
\end{lstlisting}
\centering\small{\bx}
\end{minipage}
%
\caption{
%{\small %\textmd{
\ax \CEU interface with customized \VM.
%
\bx The routine \code{VM.out} redirects all output events to the corresponding 
OS calls (ln.  1--8).
%
Each \emph{TinyOS} event callback calls \code{VM.enqueue} for the corresponding 
input event (ln 10--14).
%
System calls use \code{VM.push} for immediate return values (ln. 16--22).
%
\label{lst.terra.defs}
}
\end{figure}
